---
title: "US Retail Sales Dashboard"
author: "Elizabeth Van Der Schaaf"
date: "2025-02-23"
output: html_document
---

### Background

The data for this dashboard was taken from the Monthly Trade Report complied by the US Census (US Census Bureau, 2025). It contains retail sales data going back to 1992 broken down by many different categories. I chose this data because I thought it would yield some interesting insights into American spending, especially around the role of e-commerce and the COVID pandemic.

I want to tell the story of what Americans are spending their money on, the impact of the pandemic, and how spending patterns have changed over the years. I can use the detailed sales data broken down by month, year, and category to draw out trends or patterns in spending. My intended audience is anyone in the retail sector, but especially those interested in trends associated with e-commerce. These may be people considering a shift from physical retail or expanding their operations.

This audience is interested in any major shifts in e-commerce and physical retail. They also would want to know which retail categories Americans tend to spend more on and any current trends. They may already have a general idea of some trends but may not have taken a closer look at the numbers. They may also not be aware of sales figures for certain retail categories and how these compare to overall retail spending.

The big idea of my dashboard is that e-commerce took off during the pandemic and shows no signs of slowing down while other sectors show different behavior, notably the food service sector which showed a major decline but is now recovering.

This dashboard contains a grouped bar chart showing performance of select categories as a percentage of total retail sales. The values can be filtered by year via a slider underneath. There is also a line chart showing sales over time for certain categories. A couple of annotations were added to enhance understanding of the data. There is also the capability to filter the data using a range selector. With these filters the user is able to drill down and explore the data more. A link to the data source is included at the bottom.

### Reference
US Census Bureau. (2025, February 14). *Monthly Retail Trade* - Sales report. https://www.census.gov/retail/sales.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
library(shiny)
```

```{r}
df <- read_csv("us_retail.csv")
df$date <- mdy(df$date)
df <- arrange(df, date)

select_cats <- c("e-shopping","department","grocery","clothing","food_service")
eshopping <- filter(df, category == "e-shopping")
department <- filter(df, category == "department")
grocery <- filter(df, category == "grocery")
clothing <- filter(df, category == "clothing")
food_service <- filter(df, category == "food_service")
cats <- list(department,grocery,clothing, food_service)

# Filter data based on selected categories
selected_categories <- c("e-shopping", "department", "grocery", "clothing", "food_service")
filtered_data <- df |>
  filter(category %in% selected_categories)

# Group by category and year, and calculate total sales for each group
sales_per_category_year <- filtered_data |>
  group_by(category, year) |>
  summarise(sales = sum(sales)) |>
  ungroup()

# Do the same for total retail sales
retail_total_data <- df |>
  filter(category == 'retail_total')

retail_total_per_year <- retail_total_data |>
  group_by(year) |>
  summarise(sales = sum(sales)) |>
  ungroup()

color_map <- c("e-shopping" = "chocolate",
               "department" = "cornflowerblue",
               "grocery" = "lightsteelblue",
               "clothing" = "mediumpurple",
               "food_service" = "mediumturquoise")
```


```{r}
# Define the UI for the app
ui <- fluidPage(
  
    # Add custom CSS
  tags$style(HTML("
    .shiny-title {
      background-color: #63635e;  
      color: #faf9ed;             
      padding: 15px;
      border-radius: 4px;
    }
  ")),
  
  # Dashboard title
   titlePanel(
    div(class = "shiny-title", "US Retail Sales | The Impact of COVID 19")
  ),
  
  # Display the dynamic bar chart
  plotlyOutput("filtered_bars"),
  
  # Slider input for year selection
  sliderInput("year_slider", 
              label = "Filter bar chart by select year", 
              min = 2015, max = 2024, value = 2020, 
              step = 1,animate = TRUE, sep = ""),
  
  # Display line chart
    div(
    style = "padding-bottom: 5px;",
    plotlyOutput("line_chart")
  ),
  
    # Slider input for year range selection
  div(
    style = "padding-bottom: 50px;",
    sliderInput("range_slider", label = "Filter line chart by desired range",
              min = 2015, max = 2024, value = c(2018, 2021),
              step = 1, animate = TRUE,ticks = TRUE, sep = "")
  ),
  
  # External link to data source
  tags$div(style = "position: fixed; bottom: 10px; width: 100%; text-align: left; padding: 10px; background-color: white;",
           HTML('<a href="https://www.census.gov/retail/sales.html" target="_blank" style="font-size: 14px; color: #007bff; text-decoration: underline;">Click here for data source: Monthly Retail Trade Report</a>'))
)

  
  # Define server logic
server <- function(input, output) {
  
  output$filtered_bars <- renderPlotly({
    
    selected_year <- input$year_slider
    
    # Filter and process data
    sales <- sales_per_category_year |> 
      filter(year == selected_year)
    
    merged_data <- sales |>
      left_join(retail_total_per_year, by = "year") |>
      mutate(percentage_of_total = sales.x / sales.y) |>
      arrange(desc(percentage_of_total))
    
    plot_ly() |>
      add_trace(
        type = 'bar',
        x = merged_data$category,
        y = merged_data$percentage_of_total,
        name = 'Sales as Percentage of Total Retail Sales',
        marker = list(color = color_map[merged_data$category]),
        text = merged_data$category, 
        textposition = 'outside',
        hoverinfo = 'x+y'
      ) |>
      layout(
        title = list(text = "Sales as a Percentage of Total Retail Sales",
                     font = list(size = 22),
                     x = 0.005, xanchor = 'left',
                     y = 0.98),
        xaxis = list(
          title = selected_year,
          titlefont = list(size = 16),
          tickvals = NULL,
          ticktext = NULL,
          showline = FALSE,
          linecolor = 'grey',
          categoryorder = "array",
          categoryarray = merged_data$category
        ),
        yaxis = list(
          title = "Percent of Total Retail Sales",
          tickformat = ".0%",
          titlefont = list(size = 16),
          tickfont = list(size = 12),
          showgrid = FALSE,
          showline = FALSE,
          zeroline = FALSE,
          linecolor = 'grey'
        ),
        showlegend = FALSE,
        plot_bgcolor = 'white',
        paper_bgcolor = 'white'
      ) |>
      config(displayModeBar = FALSE)
  })
  
    # Reactive expression to filter data based on slider input
  filtered_data <- reactive({
    # Filter the data based on the selected range from the slider
    eshopping_filtered <- filter(eshopping, year(date) >= input$range_slider[1] & year(date) <= input$range_slider[2])
    department_filtered <- filter(department, year(date) >= input$range_slider[1] & year(date) <= input$range_slider[2])
    grocery_filtered <- filter(grocery, year(date) >= input$range_slider[1] & year(date) <= input$range_slider[2])
    clothing_filtered <- filter(clothing, year(date) >= input$range_slider[1] & year(date) <= input$range_slider[2])
    foodservice_filtered <- filter(food_service, year(date) >= input$range_slider[1] & year(date) <= input$range_slider[2])
    
    list(eshopping_filtered, department_filtered, grocery_filtered, clothing_filtered, foodservice_filtered)
  })
    # Create the plot based on the filtered data
  output$line_chart <- renderPlotly({
    data <- filtered_data()
    
    # Start with the e-shopping category
    line_chart <- plot_ly(data = data[[1]],
                          x = ~date,
                          y = ~sales,
                          name = "e-shopping",
                          type = 'scatter',
                          mode = 'lines',
                          line = list(color = color_map['e-shopping']))
    
    # Manually add "food_service" before the loop to ensure it's included
line_chart <- add_trace(line_chart, data = data[[5]],
                        x = ~date,
                        y = ~sales,
                        name = "food_service",
                        type = 'scatter',
                        mode = 'lines',
                        line = list(color = color_map['food_service']))
    
    # Add the traces for the other categories
    for (cat in data[2:length(data)]) {
      line_chart <- add_trace(line_chart, data = cat,
                              x = ~date,
                              y = ~sales,
                              name = unique(cat$category),
                              type = 'scatter',
                              mode = 'lines',
                              line = list(color = color_map[unique(cat$category)]))
    }
    
    # Dynamic annotations: Adjust the 'y' value based on the last sales value in the filtered data
    eshop_ann <- list(
      xref = 'paper',
      yref = 'y',
      x = 0.99,
      y = data[[1]]$sales[nrow(data[[1]])] + 2000,  # Adjust to the last value in filtered data
      xanchor = 'right',
      yanchor = 'middle',
      text = 'e-shopping',
      font = list(family = 'Arial', size = 14, color = 'rgba(67,67,67,1)'),
      showarrow = FALSE
    )

    dept_ann <- list(
      xref = 'paper',
      yref = 'y',
      x = 0.99,
      y = data[[2]]$sales[nrow(data[[2]])] + 2000,  # Adjust dynamically based on filtered data
      xanchor = 'right',
      yanchor = 'middle',
      text = 'department',
      font = list(family = 'Arial', size = 14, color = 'rgba(67,67,67,1)'),
      showarrow = FALSE
    )

    groc_ann <- list(
      xref = 'paper',
      yref = 'y',
      x = 0.99,
      y = data[[3]]$sales[nrow(data[[3]])] + 2000,  # Adjust dynamically based on filtered data
      xanchor = 'right',
      yanchor = 'middle',
      text = 'grocery',
      font = list(family = 'Arial', size = 14, color = 'rgba(67,67,67,1)'),
      showarrow = FALSE
    )

    cloth_ann <- list(
      xref = 'paper',
      yref = 'y',
      x = 0.99,
      y = data[[4]]$sales[nrow(data[[4]])] + 2000,  # Adjust dynamically based on filtered data
      xanchor = 'right',
      yanchor = 'middle',
      text = 'clothing',
      font = list(family = 'Arial', size = 14, color = 'rgba(67,67,67,1)'),
      showarrow = FALSE
    )
    
    fservice_ann <- list(
      xref = 'paper',
      yref = 'y',
      x = 0.99,
      y = data[[5]]$sales[nrow(data[[5]])] + 2000,
      xanchor = 'right',
      yanchor = 'middle',
      text = 'food service',
      font = list(family = 'Arial', size = 14, color = 'rgba(67,67,67,1)'),
      showarrow = FALSE
    )
    
    # Determine if 2020 is in the selected range
    is_2020_in_range <- 2020 >= input$range_slider[1] & 2020 <= input$range_slider[2]
    
      fservice2020_ann <- if (is_2020_in_range) {
        list(
          xref = 'x',
          yref = 'y',
          x = '2020-04-01',
          y = 31394,
          xanchor = 'left',
          yanchor = 'middle',
          text = '    From February to April<br>    food service sales declined dramatically',
          font = list(family = 'Arial', size = 12, color = 'rgba(67,67,67,1)'),
          showarrow = FALSE,
          align = 'left'
       )
      } else {
        NULL
      }
      
        eshopping2020_ann <- if (is_2020_in_range) {
        list(
          xref = 'x',
          yref = 'y',
          x = '2020-05-01',
          y = 78235,
          xanchor = 'right',
          yanchor = 'middle',
          text = 'E-commerce soared<br>between February and May    ',
          font = list(family = 'Arial', size = 12, color = 'rgba(67,67,67,1)'),
          showarrow = FALSE,
          align = 'left'
       )
      } else {
        NULL
      }
    
    # Layout for the plot
    line_chart <- line_chart |> 
      layout(
        title = list(text = "Retail Sales Over Time: The Rise of E-commerce",
                     font = list(size = 22),
                     x = 0.005, xanchor = 'left',
                     y = 0.98),
        xaxis = list(title = "", showgrid = FALSE),
        yaxis = list(title = "Sales",
                     titlecolor = 'rgba(169,169,169,1)',
                     tickformat = "$,",
                     showgrid = FALSE,
                     showline = FALSE,
                     zeroline = TRUE,
                     zerolinecolor = 'rgba(169,169,169,1)'),
        showlegend = FALSE,
        annotations = list(eshop_ann, dept_ann, groc_ann, cloth_ann, fservice_ann, fservice2020_ann, eshopping2020_ann)
      )
    
    line_chart
  })
}
# Run the Shiny app
shinyApp(ui = ui, server = server)
```

